---
title: "MLE"
output: html_notebook
---


**Maximum likelihood estimation**

To estimate a parameter, we treat the data as given and vary the parameter to find that value for which the probability of obtaining the data is highest. This value is the maximum likelihood estimate of the parameter. 

The likelihood function is also used to obtain a likelihood-based confidence interval for the parameter. This confidence interval is a large-sample approximation (i.e. frequentists assumptions about resampling), and may not be as accurate for small sample sizes.


**Log-likelihood ratio test**

Becuase of the law of likelihoods, the log-likelihood ratio test can be used to compare the fits of two nested (nested means that one model can be reduced from another model by setting one parameter = 0) models to the **same data**. 

The “full” model fits the data using the maximum likelihood estimates for the parameter(s) of interest (for example, a proportion p). 

The “reduced” model constrains the parameter values to represent a null hypothesis (for example, that p = 0.5). The G statistic is calculated as twice the difference between the log-likelihoods of the two models (“full” minus “reduced”):

`G <- 2 *(loglikefull - loglikereduced)`

G is also known as the deviance. Under the null hypothesis, G has an approximate χ2 distribution with degrees of freedom equal to the difference between the “full” and “reduced” models in the number of parameters estimated from data. We’ll work through an example below.

We’ll start by getting familiar with the commands in R to calculate probabilities.

The probability of heads in a coin toss is 0.5. If you flip a coin 10 times, what is the probability of obtaining exactly 5 heads and 5 tails?
```{r}
dbinom(5, size=10, p=0.5)
```
The fraction of human babies born who are boys is about 0.512. If 20 newborn babies are randomly sampled, what is the probability that exactly 10 are boys?
```{r}
dbinom(10, size=20, p=0.512)
```

Plot the entire probability distribution for the number of boys in families having six children. Assume the probability that any one child is a boy is 0.512.

```{r}
z <- dbinom(0:6, size=6, p=0.512)
names(z) <- as.character(0:6)
barplot(z, space=0, ylab="Probability", col = "firebrick", las = 1, xlab = "Number of boys")
```

**Challenge**
Create the above plot using ggplot
```{r}

```


**Left-handed flowers**
*An example modified from D. Schluter)

Individuals of most plant species are hermaphrodites (with both male and female sexual organs) and are therefore prone to inbreeding of the worst sort: having sex with themselves. The mud plantain, Heteranthera multiflora, has a simple mechanism to avoid such “selfing.” The style deflects to the left in some individuals and to the right in others. The anther is on the opposite side. Bees visiting a left-handed plant are dusted with pollen on their right side, which then is deposited on the styles of only right-handed plants visited later. 

To investigate the genetics of this variation, Jesson and Barrett (2002, Proc. Roy. Soc. Lond., Ser. B, Biol. Sci. 269: 1835-1839) crossed pure strains of left- and right-handed flowers, yielding only right-handed F1 offspring, which were then crossed with one another. Six of the resulting F2 offspring were left-handed, and 21 were right-handed. The expectation under a simple model of inheritance would be that their F2 offspring should consist of left- and right-handed individuals in a 1:3 ratio (i.e., 1/4 of the plants should be left-handed).

1. Generate a vector that includes a range of possible values for the population proportion of left-handed flowers, p, from 0.01 to 0.99 in increments of 0.01.

```{r}
p <- seq(0.01, 0.99, by = 0.01)
```

2. Given the results above, calculate the log-likelihood of each value for p in the F2 generation.

```{r}
loglike <- dbinom(6, size = 27, prob = p, log = TRUE)
```

3. Create a line plot of the log-likelihood against the range of values for p. 
```{r}

plot(loglike ~ p, xlab="Population proportion, p", ylab="Log-likelihood", type="l")
```

**Challenges (4, 5 and 6)**
4. Can you redraw that plot using ggplot? What is the resulting curve called? Can you see approximately the value of p corresponding to the highest point of the curve? What is this value called?


5. To get closer to this value, repeat steps (1) to (3) using a narrower range of values for p surrounding the highest point in the curve and an additional decimal point. Use your results to determine the maximum likelihood estimate of the proportion of left-handed F2 flowers. 

6. Provide a likelihood-based 95% confidence interval for the population proportion.*
*Hint* `MLE 95% CIs = MLE+/- 1.92 log-Likelhood units`


